2-the-cubble-pkg

### 2023-06-14 2-cubble.Rmd
matching code update and misc
update the cubble class intro as the pkg vignette

[conceptual framework] The \code{cubble} class tackles the spatio-temporal data collected at unique fixed locations and allowed for irregularity in the temporal dimension, i.e. irregular time steps and missingness.

The \code{cubble} class (\code{cubble_df}) is an S3 class [@advr] built on the \code{tibble} class, specifically to organize spatio-temporal data. 
It uses three attributes to character the spatio-temporal data:

  * \code{key} identifies the spatial locations 
  * \code{date} identifies the temporal order, and 
  * \code{coords} identifies the longitude and latitude coordinate associated with the location.

The base class \code{cubble_df} has two specialised class \code{spatial_cubble_df} and \code{temporal_cubble_df} to arrange the spatio-temporal data in two forms, as illustrated in Figure \ref{fig:class}.

```{r class, fig.align="center", out.width = "100%", fig.cap = "Illustration of a cubble object in the long form and nested form, along with the associated attributes."}
knitr::include_graphics(here::here("figures/diagram-keynotes/diagram-keynotes.007.png"))
```

### nested cubble

A nested cubble (in the spatial form) has class \code{c("spatial_cubble_df", "cubble_df")}. It arranges spatial variables in columns and nests temporal variables in a specialised \code{ts} column. Below prints an nested cubble object using the weather station data collected in three airports. The data is a subset of a larger data \code{climate_aus} collected from National Oceanic and Atmospheric Administration (NOAA) Global Historical Climatology Network Daily (GHCND). The data include spatial variables such as station ID, longitude, latitude, elevation, station name, world meteorology organisation ID, in additional to daily temporal information, maximum and minimum temperature values and precipitation for the year 2020. The airport data is a subset of three airport stations in Melbourne for the initial 10 days of the year (the creation of a cubble object will be explained in Section \@ref(create)):

@@

```{r echo = TRUE}
cb_nested
class(cb_nested)
```

A nested cubble header prints the key and index variable name and its form. The spatial extent is given by the bounding box with CRS information (missing here). The nested cubble header also shows the name of temporal variables nested in the \code{ts} column and their variable types: \code{prcp [dbl], tmax [dbl], tmin [dbl]}.

### long cubble

A long cubble (in the temporal form) has class \code{c("temporal_cubble_df", "cubble_df")}, which expands the temporal variables into the long form and stores the spatial variables as a data attribute.

@@

```{r echo = TRUE}
cb_long
class(cb_long)
```

In the header of a long cubble object, the extent now prints the temporal range, the interval, and whether there is gaps in the data. It also prints the available spatial variables, as oppose to the temporal variables in the nested cubble. The spatial component can be printed with \code{spatial()}:

```{r}
spatial(cb_long)
```

@@

### Coercing from foregin objects

@@

Table: (\#tab:funs) Functions implemented in the \pkg{cubble} package

|Category |   Functions|
|---|----------------|
|base R | `[`, `[<-`, `names<-`, `[[`, `[[<-`| 
|tidyverse |`dplyr_row_slice`, `dplyr_col_modify`, `dplyr_reconstruct`, `select`, `mutate`, `arrange`, `filter`, `summarise`, `select`, `slice`, `rowwise`, `rename`, `bind_cols`, `group_by`, `ungroup`, `slice_*`, `*_join`, `relocate`, `type_sum` | 
|cubble | `cubble`, `make_cubble`, `new_spatial_cubble`, `new_temporal_cubble`, `make_spatial_sf`, `make_temporal_tsibble`, `face_temporal`, `face_spatial`, `unfold`, `key`, `key_vars`, `key_data`, `index`, `index_var`, `coords`, `spatial`, `fill_gaps`, `scan_gaps`, `switch_key`, `match_sites`, `match_spatial`, `match_temporal`, `geom_glyph`, `geom_glyph_box`, `geom_glyph_line`|

Table: (\#tab:methods) Methods implemented for the three \code{cubble} classes: \code{cubble\_df}, \code{spatial\_cubble\_df}, and \code{temporal\_cubble\_df}

|Class | Methods | 
| ----- | ----------------|
|`cubble_df` | `[, dplyr_col_modify, key_data, key_vars, key, names<-, print, tbl_sum`|
|`spatial_cubble_df` | `dplyr_reconstruct, dplyr_row_slice, face_spatial, face_temporal, arrange, rename, rowwise, group_by, ungroup, select, spatial, summarise, tbl_sum, unfold, update_cubble`|
|`temporal_cubble_df` | `arrange, dplyr_reconstruct, dplyr_row_slice, face_spatial, face_temporal, fill_gaps, group_by, ungroup,  rename, rowwise, scan_gaps, select, spatial, summarise, tbl_sum, unfold, update_cubble`|

<!-- methods(class = "cubble_df") |> paste0(collapse = ", ") |> stringr::str_remove_all(".cubble_df") -->
<!-- methods(class = "spatial_cubble_df") |> paste0(collapse = ", ") |> stringr::str_remove_all(".spatial_cubble_df") -->
<!-- methods(class = "temporal_cubble_df") |> paste0(collapse = ", ") |> stringr::str_remove_all(".temporal_cubble_df") -->

### dplyr and tidyr

@@

```{r face, echo = FALSE, fig.align="center", out.width = "100%", fig.cap = "An illustration of the function \\code{face\\_temporal()} and \\code{face\\_spatial()}: \\code{face\\_temporal()} converts a spatial cubble (neseted form) into a temporal cubble (long form) to focus on the temporal variables. Conversely, \\code{face\\_spatial()} transforms a temporal cubble into a spatial one to emphasize spatial variables."}
knitr::include_graphics(here::here("figures/diagram-keynotes/diagram-keynotes.002.png"))
```

@@

```{r echo = TRUE}
#ts_long %>% has_gaps()
```

>>>>

[conceptual framework] Spatio-temporal data comes various spatial and temporal characteristics and requires different data structures to wrangle: climate weather stations are recorded at fixed point location but with potential temporal data quality issue (missingness on the day); GPS data tracks unique point locations at different timestamps; satellite imageries captures snapshots of landscape at selected time. The type of spatio-temporal data cubble tackles are those collected at unique fixed locations while allowed for irregularity in the temporal dimension, like the weather station data. In the four layouts presented by the spacetime paper [@spacetime], cubble handles full space-time and sparse space-time layouts.

The cubble class is an S3 class, built on tibble, to pivot spatio-temporal data into a nested (spatial) form and a long (temporal) form. It has two subclasses:

-   a nested/ spatial cubble has class `c("spatial_cubble_df", "cubble_df")`
-   a long/ temporal cubble has class `c("temporal_cubble_df", "cubble_df")`

A nested cubble arranges spatial variables in columns and nests temporal variables in a specialised `ts` column. Below prints an nested cubble object using the weather station data collected in three airports (the creation of a cubble object will be explained in Section \@ref(create)). This toy data is a subset of a larger data \code{climate_aus} collected from Global Historical Climatology Network Daily (GHCND). The three airport stations in Melbourne are recorded with station metadata: station ID, longitude, latitude, elevation, station name, world meteorology organisation ID. The temporal variables are precipitation, maximum and minimum temperature, which can be read from the cubble header:

@@

```{r echo = TRUE}
cb_nested
```

<!-- A nested cubble header prints the key and index variable name and its form. The spatial extent is given by the bounding box with CRS information (missing here). The nested cubble header also shows the name of temporal variables nested in the \code{ts} column and their variable types: \code{prcp [dbl], tmax [dbl], tmin [dbl]}. -->

A long cubble expands the temporal variables into the long form and stores the spatial variables as a data attribute. In the header of a long cubble object, the extent now prints the temporal range, the interval, and whether there is gaps in the data. It also prints the available spatial variables, as oppose to the temporal variables in the nested cubble.

@@

```{r echo = TRUE}
cb_long
```

### The cubble attributes

```{r class, fig.align="center", out.width = "100%", fig.cap = "Illustration of a cubble object in the long form and nested form, along with the associated attributes."}
#knitr::include_graphics(here::here("figures/diagram-keynotes/diagram-keynotes.004.png"))
```

A cubble object inherits the attributes from tibble (and its subclasses): `class`, `row.names`, and `names`, in addition to three specialised attributes:

-   `key`: the spatial identifier
-   `index`: the temporal identifier
-   `coords`: a pair of ordered coordinates associated with the location

Readers known the `key` and `index` attributes from the `tsibble` package would already be familiar the two arguments. In cubble, the `key` attribute identifies the row in the nested cubble, and together with the `index` argument, identifies the row in the long cubble. Currently, cubble only supports one variable as the key and the accepted temporal class for index includes the base R class `Date`, `POSIXlt`, `POSIXct` and tsibble's `yearmonth`, `yearweek`, and `yearquarter` class.

The `coords` attribute takes an ordered pair of coordinate. It can be a unprojected pair of longitude and latitude, or a projected easting and northing values. Under the hood, the `sf` package is used to calculate the bounding box, shown in the header of a nested cubble, and other spatial operations.

The long cubble has a special attribute `spatial` to store the spatial variables: all the variables in the nested cubble, except for the `ts` column. The shortcut function are available to extract components in the attributes, for example, \code{spatial()} for extracting the tibble object for the spatial variables:

```{r echo = TRUE}
spatial(cb_long)
```

@@

### Coercing from foreign objects

@@

| Category  | Functions |
|--------|--------------------------------------------------------|
| base R    | `[`, `[[<-`, `names<-` |
| tidyverse | `dplyr_row_slice`, `dplyr_col_modify`, `dplyr_reconstruct`, `select`, `mutate`, `arrange`, `filter`, `group_by`, `ungroup`, `summarise`, `select`, `slice`, `rowwise`, `rename`, `bind_rows`, `bind_cols`, `relocate`, `type_sum`, the slice family (`slice_head`, `slice_tail`, `slice_max`, `slice_min`, `slice_sample`) and the join family (`left_join`, `right_join`, `inner_join`, `full_join`, `anti_join`, `semi_join`) |
| cubble    | `as_cubble`, `cubble`, `make_cubble`, `check_key` `face_temporal`, `face_spatial`, `unfold`, `key`, `key_vars`, `key_data`, `index`, `index_var`, `coords`, `spatial`, `match_sites`, `match_spatial`, `match_temporal`, `geom_glyph`, `geom_glyph_box`, `geom_glyph_line`, `make_spatial_sf`, `make_temporal_tsibble`, `fill_gaps`, `scan_gaps`                                                                                |

: (#tab:funs) Functions implemented in the \pkg{cubble} package

| Class                | Methods|
|------------------|------------------------------------------------------|
| `cubble_df`          | `[[<-, dplyr_col_modify, key_data, key_vars, key, print`                                                                                                                                                                                     |
| `spatial_cubble_df`  | `[, names<-,  tbl_sum, dplyr_reconstruct, dplyr_row_slice, face_spatial, face_temporal, unfold, arrange, rename, rowwise, group_by, ungroup, select, spatial, summarise, unfold, update_cubble`                                              |
| `temporal_cubble_df` | `[, names<-, tbl_sum, arrange, dplyr_reconstruct, dplyr_row_slice, face_spatial, face_temporal, unfold, fill_gaps, group_by, ungroup,  rename, rowwise, scan_gaps, select, spatial, summarise, tbl_sum, bind_rows, bind_cols, update_cubble` |

: (#tab:methods) Methods implemented for the three \code{cubble} classes: \code{cubble\_df}, \code{spatial\_cubble\_df}, and \code{temporal\_cubble\_df}

<!-- methods(class = "cubble_df") |> paste0(collapse = ", ") |> stringr::str_remove_all(".cubble_df") -->

<!-- methods(class = "spatial_cubble_df") |> paste0(collapse = ", ") |> stringr::str_remove_all(".spatial_cubble_df") -->

<!-- methods(class = "temporal_cubble_df") |> paste0(collapse = ", ") |> stringr::str_remove_all(".temporal_cubble_df") -->

### dplyr

@@

```{r face, echo = FALSE, fig.align="center", out.width = "100%", fig.cap = "An illustration of the function \\code{face\\_temporal()} and \\code{face\\_spatial()}: \\code{face\\_temporal()} converts a spatial cubble (neseted form) into a temporal cubble (long form) to focus on the temporal variables. Conversely, \\code{face\\_spatial()} transforms a temporal cubble into a spatial one to emphasize spatial variables."}
knitr::include_graphics(here::here("figures/diagram-keynotes/diagram-keynotes.001.png"))
```

@@

```{r echo = TRUE}
ts_long %>% has_gaps()
```
